---
title: "Reproducible Research: Peer Assessment 1"
author: "Julio SOLANO"
date: "Sunday, May 17, 2015"
output: html_document
---

---------------------------------------------------------------------------------------------------------------------------------------------

### Description
This document presents the results of peer Assessments 1 of course.  



### Repository
Github repository with RMarkdown source code: [link](https://github.com/juliocesarsolano/RepData_PeerAssessment1)



### Introduction
The goal of this assignment to practice skills needed for reproducible research. Specifically this assignment use R markdown to write a report that answers the questions detailed in the sections below. In the process, the single R markdown document will be processed by knitr and be transformed into an HTML file.

---------------------------------------------------------------------------------------------------------------------------------------------

#### Load required libraries
```{r libraries, message = FALSE, warning = FALSE}
library(knitr)
library(plyr)
library(data.table)
library(ggplot2)
```


---------------------------------------------------------------------------------------------------------------------------------------------

#### Prepare the R environment

Throughout this report when writing code chunks in the R markdown document, always use echo = TRUE so that someone else will be able to read the code. First, we set echo equal a TRUE and results equal a 'hold' as global options for this document.
The document was prepared with R version x64 3.1.2, RStudio Version 0.98.1087 on Win8.1 OS.

```{r setoptions}
opts_chunk$set(echo = TRUE, results = 'hold')
```


---------------------------------------------------------------------------------------------------------------------------------------------

#### Data

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day. The data for this assignment can be downloaded from the course web site: Dataset: [Activity monitoring data 52K](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip). The variables included in this dataset are:

* **steps:** Number of steps taking in a 5-minute interval (missing values are coded as NA)  
* **date:** The date on which the measurement was taken in YYYY-MM-DD format  
* **interval:** Identifier for the 5-minute interval in which measurement was taken The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.  


---------------------------------------------------------------------------------------------------------------------------------------------

#### Loading and preprocessing

Load input data file from the current R working directory. This assignment instructions request to show any code that is needed to loading and preprocessing the data, like to:

1. Load the data (i.e. > read.csv())
2. Process/transform the data (if necessary) into a format suitable for your analysis.


```{r load}
data <- read.csv('activity.csv', header = TRUE, sep = ",",
                  colClasses=c("numeric", "character", "numeric"))
```


---------------------------------------------------------------------------------------------------------------------------------------------

#### Tidy Data

We convert the date field to Date class and interval field to Factor class.

```{r tidy}
data$date <- as.Date(data$date, format = "%Y-%m-%d")
data$interval <- as.factor(data$interval)
```

```{r review}
str(data)
```

---------------------------------------------------------------------------------------------------------------------------------------------

#### What is mean total number of steps taken per day?

1. Create a new dataset ignoring missing data NA
2. Plot a histogram of the total number of steps taken each day
3. Report the mean and median total number of steps taken per day

Now here we ignore the missing values.

```{r data.ignore}
data.ignore.na <- na.omit(data)
```

```{r plothist}
# Sums step by date
stepsxday <- aggregate(steps ~ date, data, sum)
colnames(stepsxday) <- c("date","steps")
head(stepsxday)

# Plot Histogram
h <- ggplot(stepsxday, aes(x = steps))
h + geom_histogram(fill = "red", binwidth = 1000) +
    labs(title="Histogram of Steps by Day", 
             x = "Number of Steps by Day", y = "Number of times in a day(Count)") + theme_bw() 
```

The mean and median of the number of steps taken per day:

```{r meanmedian1}
steps_mean   <- mean(stepsxday$steps, na.rm=TRUE)
steps_median <- median(stepsxday$steps, na.rm=TRUE)
```

The mean is **`r format(steps_mean,digits = 8)`** and median is **`r format(steps_median,digits = 8)`**.

---------------------------------------------------------------------------------------------------------------------------------------------

#### What is the average daily activity pattern?

We calculate the aggregation of steps by intervals of 5-minutes and convert the intervals as integers and save them in a data frame called stepsxinterval.

```{r interval}
steps_per_interval <- aggregate(data$steps, by = list(interval = data$interval),
                                FUN=mean, na.rm=TRUE)
#convert to integers
steps_per_interval$interval <- as.integer(levels(steps_per_interval$interval)[steps_per_interval$interval])
colnames(steps_per_interval) <- c("interval", "steps")
```

We make the plot with the time series of the average number of steps taken (averaged across all days) versus the 5-minute intervals:

```{r maxinterval}
ggplot(steps_per_interval, aes(x=interval, y=steps)) +   
        geom_line(color="red", size=1) +  
        labs(title="Average Daily Activity Pattern", x="Interval", y="Number of steps") +  
        theme_bw()
```

---------------------------------------------------------------------------------------------------------------------------------------------

#### Imputing missing values:

##### 1. Total number of missing values:

The total number of missing values in steps can be calculated using is.na() method to check whether the value is mising or not and then summing the logical vector.

```{r missing}
missing_vals <- sum(is.na(data$steps))
```

The total number of missing values are **`r missing_vals`**.

##### 2. Strategy for filling in all of the missing values in the dataset:

To populate missing values, we choose to replace them with the mean value at the same interval across days. In most of the cases the median is a better centrality measure than mean, but in our case the total median is not much far away from total mean, and probably we can make the mean and median meets.

We create a function NAs fill (data, pervalue) which the data arguement is the data (data frame) and pervalue arguement is the steps per interval data frame.

```{r filling}
na_fill <- function(data, pervalue) {
        na_index <- which(is.na(data$steps))
        na_replace <- unlist(lapply(na_index, FUN=function(idx){
                interval = data[idx,]$interval
                pervalue[pervalue$interval == interval,]$steps
        }))
        fill_steps <- data$steps
        fill_steps[na_index] <- na_replace
        fill_steps
}

rdata_fill <- data.frame(  
        steps = na_fill(data, steps_per_interval),  
        date = data$date,  
        interval = data$interval)
str(rdata_fill)
```

We check that are there any missing values remaining or not:

```{r checkmiss}
sum(is.na(rdata_fill$steps))
```

##### 3. A histogram of the total number of steps taken each day

Now let us plot a histogram of the daily total number of steps taken, plotted with a bin interval of 1000 steps, after filling missing values.

```{r stepxday}
fill_steps_per_day <- aggregate(steps ~ date, rdata_fill, sum)
colnames(fill_steps_per_day) <- c("date","steps")

##plotting the histogram
ggplot(fill_steps_per_day, aes(x = steps)) + 
       geom_histogram(fill = "blue", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Number of times in a day(Count)") + theme_bw() 
```

```{r meanmedian2}
steps_mean_fill   <- mean(fill_steps_per_day$steps, na.rm=TRUE)
steps_median_fill <- median(fill_steps_per_day$steps, na.rm=TRUE)
```

The mean is **`r format(steps_mean_fill,digits = 8)`** and median is **`r format(steps_median_fill,digits = 8)`**.

##### Do these values differ from the estimates from the first part of the assignment?

Yes, these values do differ slightly. We see that the values after filling the data mean and median are equal.  


##### What is the impact of imputing missing data on the estimates of the total daily number of steps?

As you can see, comparing with the calculations done in the first section of this document, we observe that while the mean value remains unchanged, the median value has shifted and virtual matches to the mean.  

Since our data has shown a t-student distribution (see both histograms), it seems that the impact of imputing missing values has increase our peak, but it's not affect negatively our predictions.   


---------------------------------------------------------------------------------------------------------------------------------------------

#### Are there differences in activity patterns between weekdays and weekends?

We do this comparison with the table with filled-in missing values.
1. Augment the table with a column that indicates the day of the week
2. Subset the table into two parts - weekends (Saturday and Sunday) and weekdays (Monday through Friday).
3. Tabulate the average steps per interval for each data set.
4. Plot the two data sets side by side for comparison. 

```{r weekday}
weekdays_steps <- function(data) {
    weekdays_steps <- aggregate(data$steps, by=list(interval = data$interval),
                          FUN=mean, na.rm=T)
    # convert to integers for plotting
    weekdays_steps$interval <- 
            as.integer(levels(weekdays_steps$interval)[weekdays_steps$interval])
    colnames(weekdays_steps) <- c("interval", "steps")
    weekdays_steps
}

data_by_weekdays <- function(data) {
    data$weekday <- 
            as.factor(weekdays(data$date)) # weekdays
    weekend_data <- subset(data, weekday %in% c("Saturday","Sunday"))
    weekday_data <- subset(data, !weekday %in% c("Saturday","Sunday"))

    weekend_steps <- weekdays_steps(weekend_data)
    weekday_steps <- weekdays_steps(weekday_data)

    weekend_steps$dayofweek <- rep("weekend", nrow(weekend_steps))
    weekday_steps$dayofweek <- rep("weekday", nrow(weekday_steps))

    data_by_weekdays <- rbind(weekend_steps, weekday_steps)
    data_by_weekdays$dayofweek <- as.factor(data_by_weekdays$dayofweek)
    data_by_weekdays
}

data_weekdays <- data_by_weekdays(rdata_fill)

# Plot
ggplot(data_weekdays, aes(x=interval, y=steps)) + 
        geom_line(color="red") + 
        facet_wrap(~ dayofweek, nrow=2, ncol=1) +
        labs(x="Interval", y="Number of steps") +
        theme_bw()
```

*We can see at the graph above that activity on the weekday has the greatest peak from all steps intervals. But, we can see too that weekends activities has more peaks over a hundred than weekday. This could be due to the fact that activities on weekdays mostly follow a work related routine, where we find some more intensity activity in little a free time that the employ can made some sport. In the other hand, at weekend we can see better distribution of effort along the time.*

